% !TEX root = ../thesis.tex

\section{Monte Carlo and Sequential Monte Carlo}
\subsection{From quadrature to sampling}
% KEEP pi (later p is filtering/smoothing)
We consider the problem of computing the expected value of a test function $\varphi$ with respect to a distribution $\pi\in\mathcal P(\mathcal X)$ i.e.: $I:=\E_\pi[\varphi(X)]$. Assuming it can't be computed analytically, a general approach is to consider a quadrature rule of the form:\add{be consistent with $n$ number of points $d$ dimensions etc.}
%
\eqa{
	I\esp\approx\esp \widehat I_n &=& \sum_{i=1}^{n} w(X^{(i)}) \varphi(X^{(i)})
}
%
for some fixed points $X^{(i)}\in\mathcal X$ and corresponding weights $w(X^{(i)})\in\mathbb R_+$. 
When the number of dimensions is low, we can consider deterministic quadrature rules such as the Gauss-Hermite quadrature.\addref 
However, as the dimensionality increases, the performance of these deterministic rules becomes catastrophic even for a large number of integration points (a well-known effect related to what Bellman called the \emph{curse of dimensionality} in dynamic programming \citep{bellman57}). 
In such cases, a broadly studied approach is the Monte Carlo integration with
%
\eqa{	
	X^{(i)}\esp \simiid\esp \pi,\quad\text{and}\quad w(X^{(i)})\spe n^{-1}. 	
\label{eq:mcsampling}}
%
The strong law of large numbers then indicates that $\widehat I_n\to I$ almost surely with approximation error scaling like $\mathcal O(n^{-1/2})$ independently of the dimensionality of the problem. This is to be compared with deterministic rules which typically have approximation error scaling like $\mathcal O(n^{-\alpha/d})$ for a fixed $\alpha>0$ depending on the quadrature rule \citep{caflisch98}. The problem then becomes one of drawing iid.\ samples from $\pi$ which is often also an intractable problem. Finally, note that \eqref{eq:mcsampling} in fact defines an \emph{empirical density} $\hat \pi$ with
%
\eqa{
	\hat \pi(x) &:=& n\inv\sum_{i=1}^{N} \delta_{X^{(i)}}(x),
}
%
and computing $\hat I_{n}$ amounts to taking the expected value of $\varphi(X)$ with respect to $\hat\pi$.

%%%%%%%
\subsection{Sequential importance sampling}
%For this point, we refer to the note by \citet{doucet11} and also to \citet[chapter 3.3 and 14.3]{robert04}.
In \emph{importance sampling} (IS), samples are drawn from a \emph{proposal distribution} $q$ that is easy to sample from (e.g.: a Gaussian) and is similar to the target distribution $\pi$. 
The quadrature weights are then adjusted to reflect that the samples are not drawn from the true distribution:
%
\eqa{
	X^{(i)}\esp \simiid\esp q(\cdot), \quad\text{and}\quad w(X^{(i)})\spe {\pi(X^{(i)})\over n q(X^{(i)})}.
} 
%
Provided the support of $q$ includes that of $\pi$ i.e., $\pi(x)>0\Rightarrow q(x)>0$, the resulting \emph{importance sampling estimator} is consistent \citep[chapter 3.3]{robert04}.\check{june21apr1}

In the context of HMM in particular, we are usually interested in estimating a sequence of target distributions $\{\pi_{t}(x_{1:t})\}_{t=1}^{T}$ which admit the following factorisation structure:
%
\eqa{
	\pi_{t}(x_{1:t}) &=& \pi_{t}(x_{t}\st x_{1:t-1})\pi_{t-1}(x_{1:t-1}).\nn
}
%
This lead to the development of \emph{sequential Monte Carlo} (SMC) methods \citep[chapter 14.3]{robert04}. The principle is the same as that of importance sampling except that a different proposal is considered at every time $t$ taking into account the previous draw of particles and the evolution of the system:
%
\eqa{
	q_{t}(x_{1:t}) &=& q_{t}(x_{t}\st x_{1:t-1})q_{t-1}(x_{1:t-1}).	\nn
}
%
Following this form, new samples or \emph{particles} $X^{(i)}_t$ can be drawn from $q_{t}(x_{t}\st X^{(i)}_{1:t-1})$ and the weights corresponding to the trajectories $\{X^{(i)}_{1:t}\}$ then need to be updated by a factor $\alpha^{(i)}_{t}$ with\check{jun24}
%
\eqa{
	\alpha^{(i)}_{t} := {\pi_{t}(X^{(i)}_{1:t} )\over \pi_{t-1}(X^{(i)}_{1:t-1}) q_{t}(X^{(i)}_{t}\st X^{(i)}_{1:t-1})}.
}
%
The variance of an IS estimator is directly related to the variance of the associated importance weights. In order to counter the increase of variance induced by the sequential IS procedure, the proposal at time $t$ should be such that the variance of the update factors $\alpha_t$ is as small as possible. In particular, the \emph{optimal proposal} \citep{doucet11} keeps it at zero with
\eqa{		q^{\text{opt}}_{t}(x_{t}\st x_{1:t-1}) &:=& \pi_{t}(x_{t}\st x_{1:t-1}).	\label{optimal proposal}}
Note that "optimality" here is understood in terms of the variance of the estimator. Additionally, since we can't typically sample easily from the optimal proposal, we have to resort to approximating distributions.\check{jun24}\add{check transition density introduced} 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Particle filtering}
In the filtering problem\add{check introduced earlier, link} on a HMM, the target densities are $\pi_{t}(x_{1:t})=p(x_{1:t}\st y_{1:t})$ and their marginals. The incremental update factors are given by
\eqa{ \alpha_t(x_{1:t}) &=& {p(x_{1:t}\st y_{1:t})\over p(x_{1:t-1}\st y_{1:t-1})q_t(x_t\st x_{1:t-1})} 	\nn\\
	&=& {p(x_t,y_t\st x_{1:t-1},y_{1:t-1})p(x_{1:t-1}\st y_{1:t-1})p(y_{1:t-1})\over p(x_{1:t-1}\st y_{1:t-1})q_t(x_t\st x_{1:t-1})p(y_{1:t})}  \nn\\
	&\propto& {p(y_t\st x_t)p(x_t\st x_{t-1}) \over q_t(x_t\st x_{1:t-1})},}
where we exploited the conditional dependence structure of a HMM. Consequently, the optimal proposal is
\eqa{		q^{\text{opt}}_{t}(x_{t}\st x_{t-1}) &\propto& p(y_{t}\st x_{t})p(x_{t}\st x_{t-1})	, \label{particle filter OID}}
which could also have obtained from applying \eqref{optimal proposal}. A skeleton of a particle filter algorithm is given below.\check{jun24}\add{in code: $\mu$ is defined where?}
%
\begin{algorithm}[!h]\small
	\caption{\label{alg:particle-filter}\dblue{\emph{\small Skeleton of a particle filter algorithm}}}
	\begin{algorithmic}[1]
		\State sample $X_{1}^{(i)}\simiid q_{1}$ for $i=1,\dots,N$	%\Comment{\emph{Initialization}}
		\State compute and normalise the weights $w_{1}(X^{(i)}_{1})\propto {\mu( X^{(i)}_{1})p(y_{1}\st X^{(i)}_{1})/ q_{1}(X^{(i)}_{1})}$
%		\State resample: $\{\X^{(i)}_{1},W^{(i)}_{1}\}\rightarrow\{\overline \X^{(i)}_{1},N\inv\}$\Comment{\emph{Using some resampling scheme}}\vspace*{.2cm}
		\For{$t=2:T$}
			\State sample $X^{(i)}_{t}\sim q_{t}( \cdot \st X^{(i)}_{t-1})$ with $q_{t}\approx q_{t}^{\text{opt}}$
			\State update and normalise the weights $w^{(i)}_{t}\propto\alpha^{(i)}_{t}w^{(i)}_{t-1}$
		\EndFor\\
		\Return weighted set of particles $\{X^{(i)}_{1:T},w^{(i)}_{1:T}\}_{i=1}^{N}$
	\end{algorithmic}
\end{algorithm}
%

The complexity of the above algorithm is $\mathcal O(TN)$. Indeed, at each time step $t$,  the algorithm samples $N$ particles and computes their corresponding weights which has linear complexity in the number of particles.\check{jun24}

Since it is often difficult to sample directly from the optimal proposal. In the context of filtering, an alternative choice is the \emph{bootstrap proposal} \citep{doucet11}: 
\eqa{q^{\text{bs}}(x_{t}\st x_{t-1})&:=&p(x_{t}\st x_{t-1}),}
which is often easier to sample from. In that case, the update factor simply reduces to $\alpha^{(i)}_t \propto p(y_t\st X^{(i)}_t)$. Since the likelihood and the transition density may not be well aligned, those update factors can vary a lot with resulting in an increase in the variance of the resulting estimator.\check{jun24}

Another important point to note is that we have omitted the \emph{resampling step} in the algorithm \ref{alg:particle-filter}. That step resamples particles with replacement based on their weights. In practice, this alleviates the problem of variance growth over time when one considers a suboptimal proposal distribution such as the bootstrap proposal. Although this step is a key aspect of particle filters, it is not an aspect that plays an important role in the issues we consider in this document; in the sequel we will assume that a standard multinomial resampling is applied \citep{doucet11}.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sampling with piecewise deterministic Markov processes}


\section{Discussion}

\todofr{
	Connect the two background sections with main element (PF, PS link to HMM, PDMP, Local PDMP for anything. Exact methods.).
}


